---
title: "Weighting and log-PPMR distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighting and log-PPMR distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mizerStomach)
```

This article explains the weighting convention used by `fit_log_ppmr()` and
`transform_fit()`. It is the conceptual reference for the shorter discussion
in `vignette("mizerStomach")`. The following article,
`vignette("feeding_kernels")`, applies the same convention to a mizer feeding
kernel.

## From prey mass to log PPMR

Let $w'$ be predator mass and $w$ prey mass. Define the predator/prey mass
ratio and its logarithm by

$$
r = \frac{w'}{w},
\qquad
l = \log(r) = \log\left(\frac{w'}{w}\right).
$$

Suppose $f_w(w\mid w')$ is the probability density of prey mass in stomachs
of predators of mass $w'$. Since $w = w'e^{-l}$ and
$|dw/dl| = w$, the density on the log-PPMR scale is

$$
f_0(l\mid w') = w f_w(w\mid w').
$$

The subscript zero anticipates that this density gives equal weight to every
prey individual. Similarly, because $w=w'/r$,

$$
f_r(r\mid w') =
\frac{w'}{r^2}f_w\left(\frac{w'}{r}\,\middle|\,w'\right).
$$

Working with $l$ is convenient because multiplicative predator/prey ratios
become additive differences and the package's fitted families have simple
weighting transformations on this scale.

## A family of weighted densities

Rows in stomach datasets can represent more than one prey individual through
`n_prey`. At weighting exponent $q$, `fit_log_ppmr()` gives row $i$ weight

$$
n_{prey,i}w_{prey,i}^{q}.
$$

For predators of a fixed mass $w'$,
$w^q=(w')^q e^{-ql}$. The factor $(w')^q$ disappears when the conditional
density is normalized, giving

$$
f_q(l\mid w') =
\frac{e^{-ql}f_0(l\mid w')}
{\int e^{-q\tilde l}f_0(\tilde l\mid w')d\tilde l}.
$$

More generally, if a fit currently represents exponent $q_0$, changing to
$q_1$ gives

$$
f_{q_1}(l) \propto
e^{-(q_1-q_0)l}f_{q_0}(l).
$$

This is the exponential tilt implemented by `transform_fit()`. The function
changes fitted parameters; it does not revisit the stomach observations.

Several exponents have useful interpretations:

| Exponent | Interpretation |
|:--|:--|
| $q=0$ | Number distribution of identifiable prey in stomachs |
| $q=2/3$ | Diet-biomass distribution under the digestion assumption below |
| $q=1$ | Biomass distribution of identifiable prey in stomachs |
| $q=\lambda-4/3$ | Feeding-kernel representation used by `set_kernel_params()` |

These are different summaries of the same stomach observations, not competing
definitions of PPMR.

### Why diet biomass corresponds to `power = 2/3`

Suppose prey mass is digested at a rate proportional to $w^{2/3}$ because the
rate is controlled by surface area. If prey remains identifiable until a fixed
fraction of its mass has been digested, its identifiable residence time is
proportional to

$$
\frac{w}{w^{2/3}} = w^{1/3}.
$$

The number of prey of mass $w$ observed in stomachs is therefore proportional
to their ingestion rate times $w^{1/3}$. Diet biomass is ingestion rate times
$w$, so it is proportional to observed stomach numbers times $w^{2/3}$. This
motivates `power = 2 / 3`; it is an assumption about digestion, not a
universal constant.

### Pooling predator sizes

The cancellation of $(w')^q$ above occurs within a fixed predator mass.
Transforming a pooled species-level fit assumes that the conditional log-PPMR
distribution is adequately independent of predator size. If that assumption
does not hold, directly fitting the pooled observations at a new `power` can
differ from transforming an old fit. Use `plot_ppmr_violins()` to investigate
the assumption and consider stratified fits when necessary.

## Transforming fitted families

The normal, smoothly truncated-exponential, and Gaussian-mixture families are
closed under exponential tilting. Let
$a=q_1-q_0$.

### Normal distribution

For a normal density with mean $\mu$ and standard deviation $\sigma$,

$$
e^{-al}\exp\left[-\frac{(l-\mu)^2}{2\sigma^2}\right]
\propto
\exp\left[-\frac{(l-(\mu-a\sigma^2))^2}{2\sigma^2}\right].
$$

The transformed mean is $\mu-a\sigma^2$ and the standard deviation is
unchanged.

```{r normal-transform}
fit_number <- fit_log_ppmr(
  barnes_data,
  "Albacore",
  distribution = "normal",
  power = 0
)
fit_biomass <- transform_fit(fit_number, power = 1)
fit_number[c("power", "mean", "sd")]
fit_biomass[c("power", "mean", "sd")]
```

### Smoothly truncated exponential

The unnormalized density used by `dtexp()` is

$$
f(l) \propto
\frac{e^{\alpha l}}
{(1+e^{u_l(l_l-l)})(1+e^{u_r(l-l_r)})}.
$$

Multiplication by $e^{-al}$ changes only the exponential slope:

$$
\tilde\alpha=\alpha-a.
$$

The cutoff locations $l_l,l_r$ and steepnesses $u_l,u_r$ are unchanged. The
implemented density is numerically normalized over the intended log-PPMR
range $0\le l\le30$; see `?dtexp` for exact evaluation behavior.

### Gaussian mixture

For a mixture

$$
f(l)=\sum_{j=1}^k p_j
\mathcal N(l\mid\mu_j,\sigma_j),
$$

each transformed component has

$$
\tilde\mu_j=\mu_j-a\sigma_j^2,
\qquad
\tilde\sigma_j=\sigma_j,
$$

and its unnormalized mixing weight is

$$
\tilde p_j^* =
p_j\exp\left(-a\mu_j+\frac{a^2\sigma_j^2}{2}\right).
$$

The final $\tilde p_j$ are obtained by dividing these weights by their sum.
The following numerical check compares this parameter transformation with
directly tilting a density on a fine grid.

```{r mixture-transform}
mixture <- list(
  species = "example",
  distribution = "gauss_mix",
  power = 0,
  min_w_pred = 0,
  p = c(0.3, 0.7),
  mean = c(3, 7),
  sd = c(1, 1.5)
)

grid <- seq(-5, 18, length.out = 20001)
dx <- grid[2] - grid[1]
original <- get_density(grid, mixture)
numerical <- exp(-0.7 * grid) * original
numerical <- numerical / sum(numerical * dx)
analytical <- get_density(grid, transform_fit(mixture, power = 0.7))

max(abs(numerical - analytical))
```

Mixture fitting itself can still be sensitive to initialization and degenerate
components. The transformation above is exact for the parameters supplied to
it, but that does not guarantee that the original mixture was a good fit.

## Practical use

Fit directly at the weighting most relevant to the analysis, then use
`transform_fit()` to inspect the implications at other powers. A large
diagnostic discrepancy after transformation indicates that the fitted family,
the predator-size-independence assumption, or both may be inadequate.

For exact arguments and return structures, see `?fit_log_ppmr`,
`?transform_fit`, and `?get_density`. For the additional assumptions that
turn a stomach distribution into a feeding kernel, continue with
`vignette("feeding_kernels")`.
