---
title: "From stomach distributions to mizer feeding kernels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From stomach distributions to mizer feeding kernels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mizerStomach)
```

A stomach-content distribution is a realized outcome: it depends on predator
preference, available prey, ingestion, digestion, and sampling. A mizer feeding
kernel [@scott2014mizer] represents only the prey-size dependence of encounter and selection.
Connecting the two therefore requires assumptions.

This article derives the transformation used by `set_kernel_params()` and
shows how it is applied. The weighting notation is introduced in
`vignette("density_functions")`; distribution selection and diagnostics are
discussed in `vignette("PPMR_distributions")`.

## Assumptions

The package uses the following approximation:

1. The log predator/prey mass-ratio distribution is adequately independent of
   predator mass over the fitted range.
2. Stomach contents are in a statistical steady state: prey of each size enter
   stomachs through ingestion at the same average rate at which they cease to
   be identifiable through digestion.
3. The mass digestion rate scales as prey mass to the power $2/3$ [@jobling1981mathematical; @bromley1994role; @andersen1999effects].
4. The prey community experienced by the predator has number density
   proportional to $w^{-\lambda}$.
5. Other effects on detectability and stomach residence time do not introduce
   an additional systematic prey-size dependence.

These assumptions are a modeling choice, not information recovered from the
stomach data. The resulting kernel should be interpreted and tested with that
uncertainty in mind.

## Ingestion

Let $w'$ be predator mass, $w$ prey mass, and
$l=\log(w'/w)$. Write the rate density at which a predator ingests prey of
mass $w$ as

$$
I(w\mid w') =
i(w')\,\phi(w'/w)\,N_c(w).
$$

Here $\phi$ is the feeding-kernel shape, $N_c(w)$ is the number density of
available prey, and $i(w')$ contains factors that depend on predator mass but
not prey mass. In a multispecies model, $N_c$ can include species interaction
strengths:

$$
N_c(w)=\sum_j\theta_{ij}N_j(w).
$$

The historical prey field experienced by sampled predators is rarely known.
The package approximates it by a power-law resource spectrum,

$$
N_c(w)=n_c w^{-\lambda}.
$$

For a mizer model, $\lambda$ is read from
`params@resource_params$lambda`; it is not fixed at two by the package.

## Digestion and stomach residence

If prey mass is digested at a rate proportional to $w^{2/3}$ and prey remains
identifiable until a fixed fraction has been digested, identifiable residence
time scales as $w^{1/3}$. The corresponding disappearance rate is

$$
D(w)=d w^{-1/3}.
$$

Let $N_w(w\mid w')$ be the stomach prey-number density. The steady-state
balance assumption is

$$
I(w\mid w')=D(w)N_w(w\mid w').
$$

Substitution gives

$$
i(w')\phi(w'/w)n_cw^{-\lambda}
=d w^{-1/3}N_w(w\mid w'),
$$

and hence, up to factors independent of prey mass,

$$
\phi(w'/w)\propto
w^{\lambda-1/3}N_w(w\mid w').
$$

## Express the result on the log-PPMR scale

Let $f_0(l)$ be the number-weighted probability density of identifiable stomach
prey. The change of variables derived in
`vignette("density_functions")` gives

$$
f_0(l)\propto w N_w(w\mid w'),
\qquad
N_w(w\mid w')\propto \frac{f_0(l)}{w}.
$$

Therefore

$$
\phi(w'/w)
\propto w^{\lambda-4/3}f_0(l)
\propto e^{-(\lambda-4/3)l}f_0(l).
$$

In the package's weighting notation, the kernel shape is the stomach
distribution represented at

$$
q_{mizer}=\lambda-\frac{4}{3}.
$$

This is exactly the target power used internally by
`set_kernel_params()`. A fit can have been estimated at any recorded
`power`; `transform_fit()` moves it from that power to $q_{mizer}$ before
model parameters are set.

## Consequences for fitted families

Suppose a fit was estimated at exponent $q_0$ and let
$a=q_{mizer}-q_0$.

For a normal distribution,

$$
\mu_{kernel}=\mu_{fit}-a\sigma^2,
\qquad
\sigma_{kernel}=\sigma.
$$

mizer represents this as a `"lognormal"` kernel with
`beta = exp(mu_kernel)` and `sigma = sigma_kernel`.

For a smoothly truncated exponential,

$$
\alpha_{kernel}=\alpha_{fit}-a,
$$

while the two cutoff locations and steepnesses are unchanged. mizer represents
this as a `"power_law"` kernel using its five `kernel_*` columns.

For a Gaussian mixture, the component means and proportions are transformed as
described in `vignette("density_functions")`. mizer represents the result as a
`"gaussian_mixture"` kernel, storing the transformed component vectors in the
list-columns `kernel_p`, `kernel_mean`, and `kernel_sd`.

## Worked mizer example

We fit the Atlantic cod diet distribution at `power = 2/3`. The North Sea
example model calls this species `"Cod"`, so the fit must be renamed before it
is transferred.

```{r fit}
cod_fit <- fit_log_ppmr(
  barnes_data,
  species = "Atlantic cod",
  distribution = "normal",
  power = 2 / 3
)
cod_fit
```

The model's resource exponent determines the exact kernel weighting:

```{r kernel-power}
params <- mizer::NS_params
kernel_power <- params@resource_params$lambda - 4 / 3
kernel_power
transform_fit(cod_fit, power = kernel_power)
```

`set_kernel_params()` performs that transformation itself. Only species
present in the supplied fit are changed, and the returned object must be
assigned.

```{r set-kernel}
model_cod_fit <- cod_fit
model_cod_fit$species <- "Cod"
rownames(model_cod_fit) <- "Cod"

params <- set_kernel_params(params, model_cod_fit)
subset(
  mizer::species_params(params),
  species == "Cod",
  select = c(species, pred_kernel_type, beta, sigma)
)
```

## Extract and inspect existing kernels

`extract_fit()` reverses the parameter mapping and returns distributions at
`power = 0`. This allows existing model kernels to be inspected with
`plot_log_ppmr_fit()`, evaluated with `get_density()`, or transformed to
another weighting.

```{r extract}
model_fits <- extract_fit(params)
model_fits["Cod", c("species", "distribution", "power", "mean", "sd")]

cod_diet_again <- transform_fit(
  model_fits["Cod", , drop = FALSE],
  power = 2 / 3
)
cod_diet_again[c("power", "mean", "sd")]
```

`extract_fit()` accepts `species_dict` when model and stomach-data names need
to be reconciled in the opposite direction.

```{r species-dictionary}
cod_from_model <- extract_fit(
  params,
  species_dict = c(Cod = "Atlantic cod")
)
cod_from_model["Atlantic cod", c("species", "distribution", "power")]
```

## Interpretation and sensitivity

The fitted kernel is conditional on the assumed prey spectrum and digestion
scaling. In particular:

- changing $\lambda$ changes the exponential tilt and therefore the kernel
  mean or slope;
- size-dependent prey availability can be mistaken for predator preference;
- digestion and identification rates may not follow one allometry for all prey
  types;
- pooled stomach observations can mix different prey fields and predator
  behaviors; and
- a good fit to stomach densities does not validate the resulting population
  dynamics.

Treat the transferred parameters as empirically informed starting values.
Evaluate them with the other evidence available for the predator and with
mizer model diagnostics or calibration. Exact parameter mappings and error
conditions are documented in `?set_kernel_params` and `?extract_fit`.

## References
