---
title: "Choosing and diagnosing a log-PPMR distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing and diagnosing a log-PPMR distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r setup}
library(mizerStomach)
```

This article develops the diagnostic choices introduced in
`vignette("mizerStomach")`. It uses only `barnes_data`, which is distributed
with the package, and focuses on questions that should be answered before a fit
is transferred to a mizer model:

1. Is one log predator/prey mass-ratio (log-PPMR) distribution adequate across
   predator sizes?
2. Which observations should receive most weight?
3. Does a fitted family describe both the relevant centre and the consequential
   tails?

The mathematical relationship among weighting exponents is derived in
`vignette("density_functions")`.

## Start with the observations

Each row of `barnes_data` describes one prey item, so `n_prey` is one. Other
datasets can aggregate repeated observations by placing their count or
frequency in `n_prey`.

```{r data}
head(barnes_data, 4)
cat(length(unique(barnes_data$species)), "predator species\n")
```

The package deliberately ignores prey identity when fitting PPMR
distributions. Consequently, a species-level fit averages over prey
communities, seasons, locations, cruises, and individual stomachs represented
in the supplied data. Decide whether that aggregation is defensible before
interpreting the result as a stable predator trait.

Common data problems include:

- rounded or imputed predator and prey masses;
- small prey that are missed, highly digested, or identified only to a broad
  taxonomic group;
- rare large prey that are absent from a finite sample;
- unequal sampling effort among predator sizes, places, or seasons; and
- rows from the same stomach or cruise that are not statistically independent.

`validate_ppmr_data()` checks the package's data contract, but these
scientific issues require knowledge of the sampling programme.

## Check dependence on predator size

A species-level fit assumes that the conditional log-PPMR distribution is
adequately stable across predator sizes. `plot_ppmr_violins()` divides the
observations into ten predator-mass groups with approximately equal numbers of
rows. The number-weighted view emphasizes small prey, while `power = 2/3`
emphasizes the part of the distribution relevant to diet biomass under the
package's digestion assumption.

```{r violins-number}
set.seed(1)
plot_ppmr_violins(barnes_data, "Atlantic cod", power = 0)
```

```{r violins-diet}
set.seed(1)
plot_ppmr_violins(barnes_data, "Atlantic cod", power = 2 / 3)
```

Look for systematic movement of the centre, spread, or shape across predator
groups. If it is substantial, options include:

- fitting a biologically justified predator-size range with `min_w_pred`;
- stratifying the data and fitting separate distributions; or
- using a hierarchical model that represents predator size and sampling
  structure explicitly.

The random perturbation used to separate tied predator masses is why the
examples set a seed.

## Choose the fitted weighting

`fit_log_ppmr()` weights row $i$ by
$n_{prey,i}w_{prey,i}^{q}$, where $q$ is `power`. A direct fit at
`power = 0` emphasizes the numerous small prey in the right tail of log PPMR.
A biomass fit at `power = 1` emphasizes the rarer large prey in the left tail.
Here we use `power = 2/3` to focus on digestion-adjusted diet biomass.

This choice is part of the estimand. It should not be selected solely because
it makes a particular parametric family look attractive.

## Normal distribution

A normal fit uses the weighted mean and weighted maximum-likelihood standard
deviation. It is easy to interpret, and maps to mizer's lognormal feeding
kernel, but it is symmetric and has unbounded tails.

```{r cod-normal}
cod_normal <- fit_log_ppmr(
  barnes_data,
  species = "Atlantic cod",
  distribution = "normal",
  power = 2 / 3
)
cod_normal
```

```{r cod-normal-plot}
plot_log_ppmr_fit(barnes_data, cod_normal, type = "histogram")
```

The fitted curve is shown at both `power = 0` and `power = 1`. A normal fit
can match the density near the fitted weighting while implying too much mass
in a transformed tail. Negative log PPMR corresponds to prey heavier than the
predator, so a curve placing substantial density there deserves particular
scrutiny.

## Smoothly truncated exponential

The truncated-exponential family has an exponential interior and smooth
logistic cutoffs at both ends. The five parameters control the slope, cutoff
locations, and cutoff steepnesses. They are estimated by weighted maximum
likelihood.

```{r cod-truncated}
cod_truncated <- fit_log_ppmr(
  barnes_data,
  species = "Atlantic cod",
  distribution = "trunc_exp",
  power = 2 / 3
)
cod_truncated
```

```{r cod-truncated-plot}
plot_log_ppmr_fit(barnes_data, cod_truncated, type = "histogram")
```

This family is useful when an unconstrained normal tail is biologically
implausible or when the observed density has a long approximately exponential
region. It is also a nonlinear five-parameter fit: inspect its curve, try
sensible starting values when necessary, and do not interpret the smallest and
largest observed prey as known physiological boundaries.

A previous truncated-exponential fit can supply starting values:

```{r refit, eval = FALSE}
cod_truncated <- fit_log_ppmr(
  barnes_data,
  distribution = "trunc_exp",
  fits = cod_truncated
)
```

## Compare several species

Fits are independent by species even when requested in one call. The following
comparison shows why one family need not be appropriate for every predator.

```{r multispecies-normal, fig.height = 7}
selected_species <- c(
  "Albacore", "Atlantic cod", "Silver hake", "Yellowfin tuna"
)

normal_fits <- fit_log_ppmr(
  barnes_data,
  species = selected_species,
  distribution = "normal",
  power = 2 / 3
)
plot_log_ppmr_fit(barnes_data, normal_fits)
```

```{r multispecies-truncated, fig.height = 7}
truncated_fits <- fit_log_ppmr(
  barnes_data,
  species = selected_species,
  distribution = "trunc_exp",
  power = 2 / 3
)
plot_log_ppmr_fit(barnes_data, truncated_fits)
```

The plots are diagnostics rather than an automatic model-selection procedure.
The package does not currently return comparable AIC values or perform a
goodness-of-fit test. Prefer the simplest family that captures the region and
tails relevant to the intended model use, and report the choice and weighting.

## Direct fitting versus transformation

Closed-form transformations describe what a fitted family implies at another
weighting. They do not guarantee that a direct fit to the newly weighted
observations has the same parameters when the family is only approximate or
when predator size affects PPMR.

```{r direct-versus-transformed}
cod_number_normal <- fit_log_ppmr(
  barnes_data, "Atlantic cod", "normal", power = 0
)
cod_diet_from_number <- transform_fit(cod_number_normal, power = 2 / 3)

rbind(
  direct_diet_fit = cod_normal[c("power", "mean", "sd")],
  transformed_number_fit =
    cod_diet_from_number[c("power", "mean", "sd")]
)
```

A large difference is useful diagnostic evidence. Fit directly at the
weighting that matters most, and use transformations to examine the
consequences elsewhere.

## Gaussian mixtures

`distribution = "gauss_mix"` can represent multimodality and uses weighted EM
updates from a deterministic initialization. Mixture likelihoods can have
local optima or nearly degenerate components, so visually inspect the result
and consider multiple initializations in a more specialized analysis.

Gaussian mixtures are supported by `get_density()` and `transform_fit()` but
cannot currently be transferred to mizer with `set_kernel_params()`. They are
therefore most useful for exploratory diagnosis rather than the final kernel
workflow.

## Interactive refinement

`fit_shiny()` can switch among families, refit the current species, and adjust
parameters while displaying the number and biomass diagnostics.

```{r shiny, eval = FALSE}
tuned <- fit_shiny(barnes_data, fits = cod_truncated)
```

Manual tuning should be recorded and justified; it is not a replacement for
checking data provenance and sampling structure.

## Reporting and limitations

For a reproducible analysis, report:

- included species and any predator-size threshold;
- mass units and how masses were measured or imputed;
- the weighting exponent;
- the fitted family and parameters;
- plots at both number and biomass weighting; and
- relevant spatial, temporal, cruise, and stomach-level clustering.

If the distribution changes among those strata, a single pooled fit estimates
a mixture of realized diets rather than a universal preference. Hierarchical or
mixed-effects models may then be more appropriate than any of the package's
simple marginal fits.

After selecting and diagnosing a family, continue with
`vignette("feeding_kernels")` to see how supported fits are transformed into
mizer feeding-kernel parameters. Exact optimizer behavior and return columns
are documented in `?fit_log_ppmr`, `?fit_truncated_exponential`, and
`?plot_log_ppmr_fit`.
